\bf Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-07-23 manuscript/main.tex
man:orthogonalization proof started

	\begin{proposition}\label{proposition:eigen_regul}
		Let
		$
		(\mathbb{B}, \mathbb{B}^{\prime \prime}, \Theta, \mathbf{y})
		$
		denote the observed data defined in \eqref{def:problem_instance}.
		At the $l$-th iteration of the PLS algorithm, the maximizer of  the squared empirical covariance defined in \eqref{def:maximizer_squared_empirical_cov_reg}, represented as the Euclidean vector form as in \eqref{def:maximizer_euclidean_form},
		is obtained by solving 
		\begin{equation}	\label{eq: Regularized generalized rayleigh quotient equation}
			\left(
			\hat{\boldsymbol{\gamma}}_1^\top, \ldots, \hat{\boldsymbol{\gamma}}_K^\top, \hat{\boldsymbol{\zeta}}^\top
			\right)^\top
			=
			\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
			\quad \text{s.t.} \quad \boldsymbol{\xi}^\top (\mathbb{B} + \Lambda \mathbb{B}^{\prime \prime}) \boldsymbol{\xi} = 1.
		\end{equation}
		where we recall that $
		V = n^{-2} (\mathbb{B} \Theta^\top \mathbf{y})(\mathbb{B} \Theta^\top \mathbf{y})^\top$ is defined in  \eqref{def:V_matrix}, and 
		$\Lambda \in \mathbb{R}^{(MK+p) \times (MK+p)}$ is given by
		\begin{equation}\label{def:Lambda}
			\Lambda := \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p}),
		\end{equation}
		where $0_{p \times p}$ denotes the $p \times p$ zero matrix.
	\end{proposition}
	The proof of Proposition \ref{proposition:eigen_regul} is provided in Appendix  \ref{section:proof:proposition:linear_regul_system_unregularized}.
	The smoothing parameter $\lambda_k > 0$ $(k = 1, \ldots, K)$ balances goodness of fit and smoothness in $\hat{\xi}_j$. Smaller $\lambda_k$ yields components that better fit the data but risks overfitting; setting $\lambda_k = 0$ recovers the unregularized solution in Proposition~\ref{proposition:eigen_noregul}. Larger $\lambda_k$ enforces greater smoothness, and in the limit $\lambda_k \to \infty$, $\hat{\xi}_j(t)$ approaches a linear form $a + bt$. In practice, both $\{\lambda_k\}$ and the number of components $L$ can be selected via cross-validation using a predictive criterion such as mean squared error.

@@

	\subsection{Step 2: Computing Residualized Predictors via Hybrid-on-Scalar Regression} \label{subsec: Step 2: Computing Predictors via Hybrid-on-Scalar Regression}
The $l$-th iteration's second step involves orthogonalization. This is achieved by fitting a hybrid-on-scalar regression model, where the predictor is the PLS score for each observation. The individual PLS score is computed as  $\hat{\rho}^{[l]}_i := \langle \widetilde{W}_i^{[l]}, \, \widehat{\xi}^{[l]} \rangle_{\mathcal{H}}$, utilizing the estimated PLS component direction $\widehat{\xi}^{[l]}$ obtained from Propositions \ref{proposition:eigen_regul} and \ref{proposition:linear_regul}.Subsequently, using the basis expansion coefficient notation from \eqref{def:truncated_zeta_and_delta}, \eqref{def:coef_vector_form}, \eqref{def:theta}, and \eqref{def:B_matrix}, the full vector of PLS scores $\hat{\boldsymbol{\rho}}^{[l]} := (\hat{\rho}^{[l]}_1, \ldots, \hat{\rho}^{[l]}_n)^\top$ is computed through the following matrix multiplication:
\begin{equation*}
( \hat{\boldsymbol{\rho}}^{[l]} )^\top = \sum_{k=1}^K ( \hat{\boldsymbol{\gamma}}_j^{[l]})^\top B \, (\Theta_j^{[l]})^\top + \boldsymbol{\zeta}^\top (\mathbf{Z}^{[l]}).
\end{equation*}
Using this as a scalar predictor vector, we fit the regression model $\widetilde{W}^{[l]}_i = \hat{\rho}^{[l]}_i\, \delta^{[l]} + E_i^{[l]}$, and then take the residuals as the predictor for the next iteration, namely $\widetilde{W}^{[l+1]}_i$. Here, $\delta^{[l]}$ and $E_i^{[l]}$ are the regression coefficient and error terms, respectively, both elements of $\tilde{\mathcal{H}}$ as defined in \eqref{def:truncated_hilbert_space}. To ensure the stability of the entire PLS estimation procedure, the estimated regression coefficient $\hat{\delta}^{[l]} = (\hat{\delta}^{[l]}_1, \ldots, \hat{\delta}^{[l]}_K, \hat{\boldsymbol{\chi}}^{[l]} )$ should have smooth functional parts, consistent with Section \ref{sec: Regularization of PLS Components}. Consequently, $\delta^{[l]}$ is obtained by minimizing a penalized least squares criterion, where the functional part is penalized by the integrated squared second derivative defined in \eqref{def:roughness_penalty}.
 \begin{align*}
\operatorname{PENSSE}( \delta)
&=
\sum_{i=1}^n \langle \widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta,
\,
\widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta
\rangle_{\mathcal{H}}
+
\tau
\sum_{j=1}^K\operatorname{PEN}(\delta_j) 
%%%%%%%%%%%%%%%%%%%%
\\&=
\sum_{i=1}^n 
\langle \widetilde{W}_i^{[l]},  \, \widetilde{W}_i^{[l]} \rangle_{\mathcal{H}}
-
\sum_{i=1}^n 
 \hat{\rho}_i^{[l]} 
\langle \delta,  \, \widetilde{W}_i^{[l]} \rangle_{\mathcal{H}}
+
\bigl(
\sum_{i=1}^n 
( \hat{\rho}_i^{[l]} )^2
\bigr)
\langle \delta,  \, \delta \rangle_{\mathcal{H}}
+
\tau
\sum_{j=1}^K\operatorname{PEN}(\delta_j) 
\\&=
 \operatorname{tr}(C^{*T} C^{*}J^{*}\Lambda) 
 - 
 2 \operatorname{tr}(d^{*}J^{*}C^{*T} \rho\Lambda) 
 +
 \rho^T \rho
  \operatorname{tr}( d^{*}J^{*}d^{*T} \Lambda) + \tau \operatorname{tr}(\ddot{d}^{*}J^{*}\ddot{d}^{*T}) \jongmin{need to be fixed}
 \end{align*}

%Using the basis expansion coefficient notation presented in \eqref{def:truncated_zeta_and_delta} and \eqref{def:coef_vector_form}, \eqref{def:theta}, \eqref{def:B_matrix}
\jongmin{Old writing starts here. I temporarily mixed the order of the paragraphs.}
%As in Section \ref{subsec: Step 1: Computing Smoothed PLS components}, the hybrid response can be expanded using basis functions as $\widetilde{W}[\mathbf{t}] = \widetilde{C}^* B^*[\mathbf{t}]$. Similarly, we expand each functional regression coefficient using the basis functions as $\pi^{(k)}(t_k) =\sum_{m=1}^M d_m^{(k)} b^{(k)}_m(t_k) = \boldsymbol{\gamma}^{(k)\top} \mathbf{b}^{(k)}(t_k)$, where $\boldsymbol{\gamma}^{(k)} = (d_1^{(k)}, \ldots, d_M^{(k)})\top \in \mathbb{R}^M$ are basis coefficients. Here, a different basis system can be flexibly chosen to expand the regression coefficient, but without loss of generality we use the same system as in Section \ref{subsec: Step 1: Computing Smoothed PLS components}. With the $(MK+p)$-dimensional row vector $\boldsymbol{\gamma}^* = (\boldsymbol{\gamma}^{(1)\top}, \ldots, \boldsymbol{\gamma}^{(K)\top}, \gamma_1, \ldots, \gamma_p)$ that stacks the basis coefficients and scalar regression coefficients, we can express the linear predictor of the regression as: 
$\widehat{\boldsymbol{\rho}} \, \boldsymbol{\delta}[\mathbf{t}]^\top = \widehat{\boldsymbol{\rho}} \, \boldsymbol{\gamma}^* B^*[\mathbf{t}]$. Finally, these basis expansions of the hybrid response and the regression coefficient enable us to re-write the hybrid-on-scalar regression model \eqref{eq: hybrid on scalar regression model} in a finite-dimensional form:
\begin{equation}
	\widetilde{C}^* B^*[\mathbf{t}] = \widehat{\boldsymbol{\rho}} \, \boldsymbol{\gamma}^* B^*[\mathbf{t}] + \mathbf{e}[\mathbf{t}].
	\label{eq: rewritten hybrid on scalar regression model}
\end{equation}
$\mathbf{e}[\mathbf{t}]$ is an $n \times (K+p)$ matrix of errors. 
\\

@@

	Again, we omit $l$ in the notations. Let $\widehat{\boldsymbol{\rho}}=(\widehat{\rho}_1, \ldots, \widehat{\rho}_n)^\top$ denote an $n$-dimensional vector of PLS scores estimated from the proposed algorithm, and let the $(K+P)$-vector $\boldsymbol{\delta}[\mathbf{t}] = [\pi(\mathbf{t})^\top, \boldsymbol{\gamma}^\top]^\top$ denote the regression coefficient consisting of the functional part $\pi(\mathbf{t}) = (\pi^{(1)}(t_1), \ldots,\pi^{(K)}(t_K))^\top \in \mathbb{R}^K$ and scalar part $\boldsymbol{\gamma} = (\gamma_1, \ldots,\gamma_p)^\top \in \mathbb{R}^p$.

@@

		Now we prove Proposition \ref{proposition:eigen_regul}.
		The penalization term 
		$\sum \limits_{j=1}^K \lambda_j 
		\operatorname{PEN}(\xi_K)$
		can be written in matrix form as:
		\begin{align*}
			\sum \limits_{j=1}^K 
			\lambda_j
			\operatorname{PEN}({\xi}_j)
			&=
			\sum \limits_{j=1}^K 
			\lambda_j
			\int_0^1 \bigl\{ \hat{\xi}_j^{\prime\prime}(t) \bigr\}^2 dt
			\\&=
			\sum \limits_{j=1}^K 
			\lambda_j
			\int_0^1 \bigl\{ \sum_{l=1}^M \theta_{ijl} \, b_l^{\prime \prime}(t) \bigr\}^2 dt
			\\&	 =
			%%%%%%%%%%%%%%%%%%%
			\sum \limits_{j=1}^K 
			\lambda_j
			\sum_{l=1}^M
			\sum_{m=1}^M
			\theta_{ijl} \theta_{ijm}
			\int_0^1
			\, b_l^{\prime \prime}(t) 
			b_m^{\prime \prime}(t)
			dt
			\\&	 =
			%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
			\sum \limits_{j=1}^K
			\lambda_j
			\boldsymbol{\gamma}_j^\top
			B^{\prime \prime} 
			\boldsymbol{\gamma}_j
			%%%%%%%%%%%%%%%%%%%%%%%%%%
			\\&= 
			\boldsymbol{\xi}^{\top} \Lambda \mathbb{B}^{\prime \prime} \boldsymbol{\xi},
		\end{align*}
		where $\boldsymbol{\xi}$ is the concatenated coefficient vector form of $\xi$ as defined in \eqref{def:maximizer_euclidean_form}, $\mathbb{B}^{\prime\prime}$ is defined in \eqref{def:gram_block}, and $\Lambda$ is defined in \eqref{def:Lambda}.

@@

>>>>

	\begin{proposition}\label{proposition:eigen_regul}
		Let
		$
		(\mathbb{B}, \mathbb{B}^{\prime \prime}, \Theta, \mathbf{y})
		$
		denote the observed data defined in \eqref{def:problem_instance}.
		At the $l$-th iteration of the PLS algorithm, the maximizer of  the squared empirical covariance defined in \eqref{def:maximizer_squared_empirical_cov_reg}, represented as the Euclidean vector form as in \eqref{def:maximizer_euclidean_form},
		is obtained by solving 
		\begin{equation}	\label{eq: Regularized generalized rayleigh quotient equation}
			\left(
			\hat{\boldsymbol{\gamma}}_1^\top, \ldots, \hat{\boldsymbol{\gamma}}_K^\top, \hat{\boldsymbol{\zeta}}^\top
			\right)^\top
			=
			\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
			\quad \text{s.t.} \quad \boldsymbol{\xi}^\top (\mathbb{B} + \Lambda \mathbb{B}^{\prime \prime}) \boldsymbol{\xi} = 1.
		\end{equation}
		where we recall that $
		V = n^{-2} (\mathbb{B} \Theta^\top \mathbf{y})(\mathbb{B} \Theta^\top \mathbf{y})^\top$ is defined in  \eqref{def:V_matrix}, and 
		$\Lambda \in \mathbb{R}^{(MK+p) \times (MK+p)}$ is given by
		\begin{equation}\label{def:Lambda}
			\Lambda := \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p}),
		\end{equation}
		where $0_{p \times p}$ denotes the $p \times p$ zero matrix.
	\end{proposition}
	The proof of Proposition \ref{proposition:eigen_regul} is provided in Appendix  \ref{section:proof:proposition:eigen_regul}.
	The smoothing parameter $\lambda_k > 0$ $(k = 1, \ldots, K)$ balances goodness of fit and smoothness in $\hat{\xi}_j$. Smaller $\lambda_k$ yields components that better fit the data but risks overfitting; setting $\lambda_k = 0$ recovers the unregularized solution in Proposition~\ref{proposition:eigen_noregul}. Larger $\lambda_k$ enforces greater smoothness, and in the limit $\lambda_k \to \infty$, $\hat{\xi}_j(t)$ approaches a linear form $a + bt$. In practice, both $\{\lambda_k\}$ and the number of components $L$ can be selected via cross-validation using a predictive criterion such as mean squared error.

@@

	\subsection{Step 2: Computing Residualized Predictors via Hybrid-on-Scalar Regression} \label{subsec: Step 2: Computing Predictors via Hybrid-on-Scalar Regression}
The $l$-th iteration's second step involves orthogonalization. This is achieved by fitting a hybrid-on-scalar regression model, where the predictor is the PLS score for each observation. The individual PLS score is computed as  $\hat{\rho}^{[l]}_i := \langle \widetilde{W}_i^{[l]}, \, \hat{\xi}^{[l]} \rangle_{\mathcal{H}}$, utilizing the estimated PLS component direction $\widehat{\xi}^{[l]}$ obtained from Propositions \ref{proposition:eigen_regul} and \ref{proposition:linear_regul}. Subsequently, using the basis expansion coefficient notation from \eqref{def:truncated_zeta_and_delta}, \eqref{def:coef_vector_form}, \eqref{def:theta}, and \eqref{def:B_matrix}, the full vector of PLS scores $\hat{\boldsymbol{\rho}}^{[l]} := (\hat{\rho}^{[l]}_1, \ldots, \hat{\rho}^{[l]}_n)^\top$ is computed through the following matrix multiplication:
\begin{equation*}
( \hat{\boldsymbol{\rho}}^{[l]} )^\top = \sum_{k=1}^K ( \hat{\boldsymbol{\gamma}}_j^{[l]})^\top B \, (\Theta_j^{[l]})^\top + \boldsymbol{\zeta}^\top (\mathbf{Z}^{[l]}).
\end{equation*}
Using this as a scalar predictor vector, we fit the regression model $\widetilde{W}^{[l]}_i = \hat{\rho}^{[l]}_i\, \delta^{[l]} + E_i^{[l]}$, and then take the residuals as the predictor for the next iteration, namely $\widetilde{W}^{[l+1]}_i$. Here, $\delta^{[l]}$ and $E_i^{[l]}$ are the regression coefficient and error terms, respectively, both elements of $\tilde{\mathcal{H}}$ as defined in \eqref{def:truncated_hilbert_space}. To ensure the stability of the entire PLS estimation procedure, the estimated regression coefficient $\hat{\delta}^{[l]} = (\hat{\delta}^{[l]}_1, \ldots, \hat{\delta}^{[l]}_K, \hat{\boldsymbol{\chi}}^{[l]} )$ should have smooth functional parts, consistent with Section \ref{sec: Regularization of PLS Components}. Consequently, $\delta^{[l]}$ is obtained by minimizing a penalized least squares criterion, where the functional part is penalized by the integrated squared second derivative defined in \eqref{def:roughness_penalty}:
\begin{equation}\label{def:argmin_pensse}
	\hat{\delta}^{[l]} := \arg \min_{\delta \in \tilde{\mathcal{H}}} \operatorname{PENSSE}( \delta),
\end{equation}
where
\begin{equation}\label{def:pensse}
	\operatorname{PENSSE}( \delta)
	:=
	\sum_{i=1}^n \langle \widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta,
	\,
	\widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta
	\rangle_{\mathcal{H}}
	+
	\tau
	\sum_{j=1}^K\operatorname{PEN}(\delta_j).
\end{equation}
The following proposition shows that the solution of the problem \ref{def:argmin_pensse} can be obtained by solving linear equations.

 \begin{proposition}[Closed-form  orthogonalization]\label{proposition:closed_form_orthgonalization}
 	asdasdasd
 \end{proposition}
 Proof of Proposition \ref{proposition:closed_form_orthgonalization} is provided in Appendix \ref{section:proof:proposition:closed_form_orthgonalization}.
%Using the basis expansion coefficient notation presented in \eqref{def:truncated_zeta_and_delta} and \eqref{def:coef_vector_form}, \eqref{def:theta}, \eqref{def:B_matrix}
\jongmin{Old writing starts here. I temporarily mixed the order of the paragraphs.}
%As in Section \ref{subsec: Step 1: Computing Smoothed PLS components}, the hybrid response can be expanded using basis functions as $\widetilde{W}[\mathbf{t}] = \widetilde{C}^* B^*[\mathbf{t}]$. Similarly, we expand each functional regression coefficient using the basis functions as $\pi^{(k)}(t_k) =\sum_{m=1}^M d_m^{(k)} b^{(k)}_m(t_k) = \boldsymbol{\gamma}^{(k)\top} \mathbf{b}^{(k)}(t_k)$, where $\boldsymbol{\gamma}^{(k)} = (d_1^{(k)}, \ldots, d_M^{(k)})\top \in \mathbb{R}^M$ are basis coefficients. Here, a different basis system can be flexibly chosen to expand the regression coefficient, but without loss of generality we use the same system as in Section \ref{subsec: Step 1: Computing Smoothed PLS components}. With the $(MK+p)$-dimensional row vector $\boldsymbol{\gamma}^* = (\boldsymbol{\gamma}^{(1)\top}, \ldots, \boldsymbol{\gamma}^{(K)\top}, \gamma_1, \ldots, \gamma_p)$ that stacks the basis coefficients and scalar regression coefficients, we can express the linear predictor of the regression as: 
$\widehat{\boldsymbol{\rho}} \, \boldsymbol{\delta}[\mathbf{t}]^\top = \widehat{\boldsymbol{\rho}} \, \boldsymbol{\gamma}^* B^*[\mathbf{t}]$. Finally, these basis expansions of the hybrid response and the regression coefficient enable us to re-write the hybrid-on-scalar regression model \eqref{eq: hybrid on scalar regression model} in a finite-dimensional form:
\begin{equation}
	\widetilde{C}^* B^*[\mathbf{t}] = \widehat{\boldsymbol{\rho}} \, \boldsymbol{\gamma}^* B^*[\mathbf{t}] + \mathbf{e}[\mathbf{t}].
	\label{eq: rewritten hybrid on scalar regression model}
\end{equation}
$\mathbf{e}[\mathbf{t}]$ is an $n \times (K+p)$ matrix of errors. 
\\

@@

@@

		\section{Proof of Proposition \ref{proposition:eigen_regul} }\label{section:proof:proposition:eigen_regul}
		The penalization term 
		$\sum \limits_{j=1}^K \lambda_j 
		\operatorname{PEN}(\xi_K)$
		can be written in matrix form as:
		\begin{align*}
			\sum \limits_{j=1}^K 
			\lambda_j
			\operatorname{PEN}({\xi}_j)
			&=
			\sum \limits_{j=1}^K 
			\lambda_j
			\int_0^1 \bigl\{ \hat{\xi}_j^{\prime\prime}(t) \bigr\}^2 dt
			\\&=
			\sum \limits_{j=1}^K 
			\lambda_j
			\int_0^1 \bigl\{ \sum_{l=1}^M \theta_{ijl} \, b_l^{\prime \prime}(t) \bigr\}^2 dt
			\\&	 =
			%%%%%%%%%%%%%%%%%%%
			\sum \limits_{j=1}^K 
			\lambda_j
			\sum_{l=1}^M
			\sum_{m=1}^M
			\theta_{ijl} \theta_{ijm}
			\int_0^1
			\, b_l^{\prime \prime}(t) 
			b_m^{\prime \prime}(t)
			dt
			\\&	 =
			%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
			\sum \limits_{j=1}^K
			\lambda_j
			\boldsymbol{\gamma}_j^\top
			B^{\prime \prime} 
			\boldsymbol{\gamma}_j
			%%%%%%%%%%%%%%%%%%%%%%%%%%
			\\&= 
			\boldsymbol{\xi}^{\top} \Lambda \mathbb{B}^{\prime \prime} \boldsymbol{\xi},
		\end{align*}
		where $\boldsymbol{\xi}$ is the concatenated coefficient vector form of $\xi$ as defined in \eqref{def:maximizer_euclidean_form}, $\mathbb{B}^{\prime\prime}$ is defined in \eqref{def:gram_block}, and $\Lambda$ is defined in \eqref{def:Lambda}.

@@

	\section{Proof of Proposition \ref{proposition:closed_form_orthgonalization} \ref{}}\label{section:proof:proposition:closed_form_orthgonalization}
	 \begin{align*}
		\operatorname{PENSSE}( \delta)
		&=
		\sum_{i=1}^n \langle \widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta,
		\,
		\widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta
		\rangle_{\mathcal{H}}
		+
		\tau
		\sum_{j=1}^K\operatorname{PEN}(\delta_j) 
		%%%%%%%%%%%%%%%%%%%%
		\\&=
		\sum_{i=1}^n 
		\langle \widetilde{W}_i^{[l]},  \, \widetilde{W}_i^{[l]} \rangle_{\mathcal{H}}
		-
		\sum_{i=1}^n 
		\hat{\rho}_i^{[l]} 
		\langle \delta,  \, \widetilde{W}_i^{[l]} \rangle_{\mathcal{H}}
		+
		\bigl(
		\sum_{i=1}^n 
		( \hat{\rho}_i^{[l]} )^2
		\bigr)
		\langle \delta,  \, \delta \rangle_{\mathcal{H}}
		+
		\tau
		\sum_{j=1}^K\operatorname{PEN}(\delta_j) 
		\\&=
		\operatorname{tr}(C^{*T} C^{*}J^{*}\Lambda) 
		- 
		2 \operatorname{tr}(d^{*}J^{*}C^{*T} \rho\Lambda) 
		+
		\rho^T \rho
		\operatorname{tr}( d^{*}J^{*}d^{*T} \Lambda) + \tau \operatorname{tr}(\ddot{d}^{*}J^{*}\ddot{d}^{*T}) \jongmin{need to be fixed}
	\end{align*}
